Liveness-Based Garbage Collection for Lazy Languages

### 2015-10-03 paper/paper.tex
Added a space
Fixed typos

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,52)(0,-52)
    %\psframe(90,48)(0,-48)
    \renewcommand{\arraystretch}{1}
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\CAR\  \pl))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ \\
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ \\
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection for
        an ideal collector.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
%  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

@@

While an ideal garbage collector is  impractical, in this work we show
that  static analysis  of  heap data  can help  practical garbage
collectors in reclaiming more  garbage.  The specific contributions of
this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language  and prove  of its correctness.  We show  how to
  compute  function  summaries  in  a context  independent  manner  to
  achieve  a  context  sensitive  analysis. We  further  describe  the
  transformation of the  liveness equations to grammar rules  in a way
  that the language  accepted by the resulting  grammar describes live
  heap locations.
\item We prove  the undecidability of the grammar  that results from
  the  analysis.   To overcome  this,  we  present an  algorithm  to
  transform grammar  to finite  state automata  (DFA) such  that the
  language accepted by the  DFA over-approximates the liveness (and
  hence a safe approximation).
\item We have implemented a garbage  collector that can use the DFA
  to  retain live  cells that  also include  closures. We  have done
  extensive  experiments with  this garbage  collector and  obtained
  empirical results  that show  the effectiveness  of liveness-based
  garbage collection.
\end{itemize}

@@

\begin{figure*}[t!]
\begin{center}\footnotesize
\renewcommand{\arraystretch}{1}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

The     start      state     is      $([\;]_\rho,([\;]_\rho,     \ans,
(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm)$.  In this, the initial
environment  and  the initial  heap  are  denoted by  $[\;]_\rho$  and
$[\;]_H$,  and the  initial stack  consists of  a single  continuation
frame in which \ans\ is a distinguished variable that would be updated
with  the value  of (\mainpgm).   In addition,  \print\ is  a function
modeling a printing mechanism---a standard run-time support assumption
for lazy  languages that prints  the value of (\mainpgm).   Notice the
use of the operator $:$ for adding elements to the top of the stack.

@@

 Since  in a  lazy language,  an  expression is  not evaluated  unless
 required, the null demand ($\emptyset$) does not generate liveness in
 any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
 non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed to  the
 liveness $\{x.\epsilon, x.\acdr\sigma\}$.  In  an opposite sense, the
 demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
 demand  $\sigma$  on $z$.   Since  \CONS\  does not  dereference  its
 arguments, there  is no $\epsilon$ demand  on $y$ and $z$.  The rules
 for (\PRIM~x~y) and (\NULLQ~x) are similar. Constants do not generate
 any liveness.

For the  case of a  function call we  use a third  parameter $\Lfonly$
that  represents  the  summaries  of all  functions  in  the  program.
$\Lfonly_{\mathit  f}$  (the component  of  $\Lfonly$  for a  specific
function $f$), expresses  how the demand $\sigma$ on a  call to $f$ is
transformed into  the liveness of  its parameters at the  beginning of
the   call.    $\Lfonly$  is   determined   by   the  judgement   form
$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
This  rule  describes the  fixed-point  property  to be  satisfied  by
$\Lfonly$, namely, the demand transformation assumed for each function
in  the  program should  be  the  same  as the  demand  transformation
calculated    from    its    body.      As    we    shall    see    in
Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
grammar and  the the language generated  by this grammar is  the least
solution satisfying  the rule. We  prefer the least solution  since it
ensures the safe collection of the greatest amount of garbage.

@@

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_6$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live,  since, during execution, \py\ is yet to
  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
  consulted by the garbage collector at this point.
\item In an  eager language, the heap location referred  by \py\ would
  not have been live at $\pi_8$,  a program point beyond its last use.
  In  a  lazy  language,   however,  the  definition  $\pz  \leftarrow
  (1+\py)$, does not  result in an evaluation of  $(1+\py)$; instead a
  closure  is  created.  The  evaluation  actually  takes place  while
  evaluating $(\SRETURN~\pz)$.   We therefore  record the  liveness of
  $\py$ at  the program point  $\pi_8$, indeed at every  program point
  from $\pi_8$ up to the beginning of the program $\pi_1$ (since we do
  not kill liveness).
\end{enumerate}

@@

Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
computes a  map that takes a  e-path in ${\mathit f}$  and returns the
liveness environment  arising out of  all uses of variables  along the
e-path.  A second map $\cal{P}_{\mathit  f}$ maps any program point in
$\mathit f$  to the set  of e-paths sharing  the point.  If  a program
point happens to fall on more  than one such e-path, then the liveness
environment of  the program  point is the  variable-wise union  of the
liveness  environments  of the  individual  e-paths.   In the  example
program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  \epath_1, \epath_2,
\epath_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(\epath_1)  \cup {\cal
  M}(\epath_2) \cup  {\cal M}(\epath_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
\sigma, \Lfonly)$.  In the  sequel we shall only define $\mathcal{L}$.
The formulation  of $\cal{P}$ is easy  and we do not  elaborate it any
further.

Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
evaluation path  terminating in  $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a point-wise-union of the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
added  to the  union.   Also notice  a  consequence of  laziness---the
entire  expression including  the condition  is not  evaluated if  the
demand on it is $\emptyset$.  This  results in the empty map, which is
denoted by $\emptyset_{\cal{M}}$.

@@

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
  on the expression $e$, and $e_f$ is the body of the function. As the
  section on liveness  analysis shows, the entire  function body $e_f$
  and $\sigma$  are needed  to calculate  the liveness  environment at
  different program  points. The  demand sigma  is ``dynamic''  in the
  sense  that it  arises from  the actual  function calls  made during
  execution.   The  0-CFA  demand  is an  over-approximation  of  this
  demand.  The information  in a suspended evaluation  context is also
  similarly augmented with  the demand on it and the  function body of
  which the suspended evaluation is a  part.  Thus a stacked entry now
  takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
  small step semantics which carries the extra information is shown in
  Figure~\ref{fig:minefield-semantics}.
\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
  e_f$ and $LF$, we can construct a liveness environment as follows:
  \begin{enumerate}
  \item If $e$ is an expression, then:  
  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
    \text{where }
  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
\item If $e$ is an application, then:  
  $$ \Lv = ref(e, \sigma, \Lfonly)$$
  \end{enumerate}
We can define in the same way the liveness environment for each of the
suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
environments that we shall denote $\mathsf{SL}$.
\item $GC$ models a liveness based garbage collection:
$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
  \in domain(\rho)$,  $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv'$.
\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \in \Lv$, .
  \end{enumerate}
\end{enumerate}

@@

We  need the  following proposition that  connects  liveness analysis  with
minefield semantics transitions. 
\begin{proposition}
\label{prop:correctness}
\begin{enumerate} 
\item   Consider  a   \LET\  definition:   \linebreak  $\pi\!\!\!:   (\LET~x
  \leftarrow~\ldots~\IN~e$, such  that there is  a use of $x$  in $e$,
  i.e.  an occurrence of $\LET~y~\leftarrow~s~in \ldots$, where $s$ is
  $(\CAR\ x)$, $(\CDR\ x)$  or $(+\ x\ y)$, or $(\SIF\  x\ e1\ e2)$ or
  $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null.
\item    Assume    that    a   \LET\    definition    $\pi:    (\LET~z
  \leftarrow~(\CONS~x~y)~\IN~e$  has an  occurrence  of $(\CAR~z)$  in
  $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
  Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of$y$  at
  $\pi$ includes $\epsilon$.
\end{enumerate}
\end{proposition}

@@

\begin{proof}
For a program  executing according to {\em minefield  semantics}, if a
\bang\ has to  occur then some variable which is  dereferenced must be
set to  $\bot$, between its  creation and  use points.  This  can only
happen if  the liveness of  the variable being dereferenced  was null.
From  the Proposition~\ref{prop:correctness}  it  is clear  that this  cannot
happen.
\end{proof}

@@

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFA are converted to DFA.

>>>>

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,52)(0,-52)
    %\psframe(90,48)(0,-48)
    \renewcommand{\arraystretch}{1}
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\length (\CDR\  \pl)))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ \\
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ \\
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection for
        an ideal collector.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
%  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

@@

While an ideal garbage collector is  impractical, in this work we show
that  static analysis  of  heap data  can help  practical garbage
collectors in reclaiming more  garbage.  The specific contributions of
this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language  and prove its correctness.  We show  how to
  compute  function  summaries  in  a context  independent  manner  to
  achieve  a  context  sensitive  analysis. We  further  describe  the
  transformation of the  liveness equations to grammar rules  in a way
  that the language  accepted by the resulting  grammar describes live
  heap locations.
\item We prove  the undecidability of the grammar  that results from
  the  analysis.   To overcome  this,  we  present an  algorithm  to
  transform grammars  to finite  state automata  (DFA) such  that the
  language accepted by the  DFA over-approximates the liveness (and
  hence a safe approximation).
\item We have implemented a garbage  collector that can use the DFA
  to  retain live  cells that  also include  closures. We  have done
  extensive  experiments with  this garbage  collector and  obtained
  empirical results  that show  the effectiveness  of liveness-based
  garbage collection.
\end{itemize}

@@

\begin{figure*}[t!]
\begin{center}\footnotesize
\renewcommand{\arraystretch}{1}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{The small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

The     start      state     is      $([\;]_\rho,([\;]_\rho,     \ans,
(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm))$.  In this, the initial
environment  and  the initial  heap  are  denoted by  $[\;]_\rho$  and
$[\;]_H$,  and the  initial stack  consists of  a single  continuation
frame in which \ans\ is a distinguished variable that would be updated
with  the value  of (\mainpgm).   In addition,  \print\ is  a function
modeling a printing mechanism---a standard run-time support assumption
for lazy  languages---that prints  the value of (\mainpgm).   Notice the
use of the operator $:$ for adding elements to the top of the stack.

@@

 Since  in a  lazy language,  an  expression is  not evaluated  unless
 required. The null demand ($\emptyset$) does not generate liveness in
 any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
 non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed to  the
 liveness $\{x.\epsilon, x.\acdr\sigma\}$.  In  an opposite sense, the
 demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
 demand  $\sigma$  on $z$.   Since  \CONS\  does not  dereference  its
 arguments, there  is no $\epsilon$ demand  on $y$ and $z$.  The rules
 for (\PRIM~x~y) and (\NULLQ~x) are similar. Constants do not generate
 any liveness.

For the  case of a  function call we  use a third  parameter $\Lfonly$
that  represents  the  summaries  of all  functions  in  the  program.
$\Lfonly_{\mathit  f}$  (the component  of  $\Lfonly$  for a  specific
function $f$), expresses  how the demand $\sigma$ on a  call to $f$ is
transformed into  the liveness of  its parameters at the  beginning of
the   call.    $\Lfonly$  is   determined   by   the  judgement 
$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
This  rule  describes the  fixed-point  property  to be  satisfied  by
$\Lfonly$, namely, the demand transformation assumed for each function
in  the  program should  be  the  same  as the  demand  transformation
calculated    from    its    body.      As    we    shall    see    in
Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
grammar and  the the language generated  by this grammar is  the least
solution satisfying  the rule. We  prefer the least solution  since it
ensures the safe collection of the greatest amount of garbage.

@@

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_6$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live,  since, during execution, \py\ is yet to
  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
  considered by the garbage collector at this point.
\item In an  eager language, the heap location referred  by \py\ would
  not have been live at $\pi_8$,  a program point beyond its last use.
  In  a  lazy  language,   however,  the  definition  $\pz  \leftarrow
  (1+\py)$, does not  result in an evaluation of  $(1+\py)$; instead a
  closure  is  created.  The  evaluation  actually  takes place  while
  evaluating $(\SRETURN~\pz)$.   We therefore  record the  liveness of
  $\py$ at  the program point  $\pi_8$, indeed at every  program point
  from $\pi_8$ up to the beginning of the program $\pi_1$ (since we do
  not kill liveness).
\end{enumerate}

@@

Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
computes a  map that takes an  e-path in ${\mathit f}$  and returns the
liveness environment  arising out of  all uses of variables  along the
e-path.  A second map $\cal{P}_{\mathit  f}$ maps any program point in
$\mathit f$  to the set  of e-paths sharing  the point.  If  a program
point happens to fall on more  than one such e-path, then the liveness
environment of  the program  point is the  variable-wise union  of the
liveness  environments  of the  individual  e-paths.   In the  example
program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  \epath_1, \epath_2,
\epath_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(\epath_1)  \cup {\cal
  M}(\epath_2) \cup  {\cal M}(\epath_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
\sigma, \Lfonly)$.  In the  sequel we shall only define $\mathcal{L}$.
The formulation  of $\cal{P}$ is easy  and we do not  elaborate it any
further.

Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
evaluation path  terminating in  $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a point-wise-union of the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
added  to the  union.   Also notice  a  consequence of  laziness: the
entire  expression including  the condition  is not  evaluated if  the
demand on it is $\emptyset$.  This  results in the empty map, which is
denoted by $\emptyset_{\cal{M}}$.

@@

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
  on the expression $e$, and $e_f$ is the body of the function. As the
  section on liveness  analysis shows, the entire  function body $e_f$
  and $\sigma$  are needed  to calculate  the liveness  environment at
  different program  points. The  demand sigma  is ``dynamic''  in the
  sense  that it  arises from  the actual  function calls  made during
  execution.   The  0-CFA  demand  is an  over-approximation  of  this
  demand.  The information  in a suspended evaluation  context is also
  similarly augmented with  the demand on it and the  function body of
  which the suspended evaluation is a  part.  Thus a stacked entry now
  takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
  small step semantics which carries the extra information is shown in
  Figure~\ref{fig:minefield-semantics}.
\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
  e_f$ and $\mathcal{LF}$, we can construct a liveness environment as follows:
  \begin{enumerate}
  \item If $e$ is an expression, then:  
  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
    \text{where }
  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
\item If $e$ is an application, then:  
  $$ \Lv = ref(e, \sigma, \Lfonly)$$
  \end{enumerate}
We can define in the same way the liveness environment for each of the
suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
environments that we shall denote $\mathsf{SL}$.
\item $GC$ models a liveness based garbage collection:
$GC(\rho,$ $ S,$ $ H,$ $ \mathcal{L},$ $ \mathsf{SL}) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
  \in domain(\rho)$,  $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv'$.
\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \in \Lv$, .
  \end{enumerate}
\end{enumerate}

@@

We  need the  following proposition that  connects  liveness analysis  with
minefield semantics transitions. 
\begin{proposition}
\label{prop:correctness}
\begin{enumerate} 
\item   Consider  a   \LET\  definition:   \linebreak  $\pi\!\!\!:   (\LET~x
  \leftarrow~\ldots~\IN~e)$, such  that there is  a use of $x$  in $e$,
  i.e.  an occurrence of $\LET~y~\leftarrow~s~in \ldots$, where $s$ is
  $(\CAR\ x)$, $(\CDR\ x)$  or $(+\ x\ y)$, or $(\SIF\  x\ e1\ e2)$ or
  $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null.
\item    Assume    that    a   \LET\    definition    $\pi:    (\LET~z
  \leftarrow~(\CONS~x~y)~\IN~e)$  has an  occurrence  of $(\CAR~z)$  in
  $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
  Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of $y$  at
  $\pi$ includes $\epsilon$.
\end{enumerate}
\end{proposition}

@@

\begin{proof}
For a program  executing according to {\em minefield  semantics}, if a
\bang\ occurs then some variable which is  dereferenced must be
set to  $\bot$, between its  creation and  use points.  This  can only
happen if  the liveness of  the variable being dereferenced  was null.
From  the Proposition~\ref{prop:correctness}  it  is clear  that this  cannot
happen.
\end{proof}

@@

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFAs are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFAs.
